Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-19 paper/sections/conclusion.rmd
incorporated Cynthia's comments

This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse. We demonstrate that long-held beliefs as to what defines optimality in AR, may not be suitable in contexts that involves large groups of individuals facing adverse outcomes. Specifically, we run experiments that simulate the application of recourse in practice using various state-of-the-art counterfactual generators and find that all of them induce substantial domain and model shifts. We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse in these types of situations. By proposing an adapted counterfactual search objective that incorporates this cost, we make that paradigm shift explicit. We show that this modified objective lends itself to mitigation strategies that can be used to effectively decrease the magnitude of induced domain and model shifts. Through our work we hope to inspire future research on this important topic. To this end we have open-sourced all of our code along with a Julia package - [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md). The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022CounterfactualExplanations]. That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.

>>>>

This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse. We demonstrate that long-held beliefs as to what defines optimality in AR, may not be suitable in contexts that involves large groups of individuals facing adverse outcomes. Specifically, we run experiments that simulate the application of recourse in practice using various state-of-the-art counterfactual generators and find that all of them induce substantial domain and model shifts. We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse in these types of situations. By proposing an adapted counterfactual search objective that incorporates this cost, we make that paradigm shift explicit. We show that this modified objective lends itself to mitigation strategies that can be used to effectively decrease the magnitude of induced domain and model shifts. Through our work we hope to inspire future research on this important topic. To this end we have open-sourced all of our code along with a Julia package: [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md). The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022counterfactualexplanations]. That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.

### 2022-08-19 paper/sections/discussion.rmd
incorporated Cynthia's comments

So, where does this leave us? We would argue that the expected external costs of individual recourse should be shared by all stakeholders. The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \@ref(eq:collective). This will on average lead to more costly counterfactual outcomes. But it may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals. We have shown various types of shift-mitigating strategies that can be used to this end. Since all of these strategies can be seen simply as specific adaption of Equation \@ref(eq:collective), they can be applied to any of the various counterfactual generators studied here.

>>>>

So, where does this leave us? We would argue that the expected external costs of individual recourse should be shared by all stakeholders. The most straightforward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \@ref(eq:collective). This will on average lead to more costly counterfactual outcomes, but may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals. We have shown various types of shift-mitigating strategies that can be used to this end. Since all of these strategies can be seen simply as specific adaption of Equation \@ref(eq:collective), they can be applied to any of the various counterfactual generators studied here.

### 2022-08-19 paper/sections/empirical.rmd
incorporated Cynthia's comments

This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\@ref(empirical-2)). For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section. A few high-level details upfront: each experiment is run for a total of $T=50$ rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so $B_t=0.05 * N_t^{\mathcal{D}_0}$^[As mentioned in the previous section, we end up providing recourse to a total of $\approx50\%$ by the end of round $T=50$.]. All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment. Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model. Evaluation metrics are computed and stored every 10 rounds. To account for noise each individual experiment is repeat five times.

## $M$ -- Classifiers and Generative Models {#empirical-classifiers}

For each dataset and generator we look at three different types of classifiers all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier - **Logistic Regression** - implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier. We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty. They are also the model of choice in @schut2021generating. The actual neural network architectures are kept simple (top half of Table \@ref(tab:architecture)), since we are only marginally concerned with achieving good initial classifier performance.

@@

## $\mathcal{D}$ -- Data {#empirical-data}

We have chosen to work with both synthetic and real-world datasets. Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics. Following @upadhyay2021towards, we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results. Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.

@@

We use four synthetic binary classification datasets consisting of 1000 samples each: **Overlapping**, **Linearly Separable**, **Circles** and **Moons**. The datasets are presented in Figure \@ref(fig:synthetic-data). Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

@@

Ex-ante we expect to see that by construction Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary. Thus, the choice of a black-box model may have an impact on the paths of the recourse. For generators that use latent space search (REVISE @joshi2019towards, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples. Of course, this is expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty. Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class^[As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated. We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.]. In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concerned with generating what we have defined as meaningful counterfactuals.

@@

We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification. Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data]. Originally consisting of 250,000 instances with 11 numerical attributes. Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment. In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been omitted from the analysis. See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding. Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019towards and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.

As a third dataset we include the **California Housing** dataset derived from the 1990 U.S. census and sourced through scikit-learn [@pedregosa2011scikit, @pace1997sparse]. It consists of 8 continuous features that can be used to predict the median house price for California districts. The continuous outcome variable is binarized as $\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}$ indicating whether or not the median house price of a given district is above or below the median of all districts. While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example @schut2021generating). While we initially also conducted experiments on that dataset, we eventually discarded this dataset, since it has been found to suffer from an ethical problem [@carlisle2019racist].

@@

## $G$ -- Generators

All generators introduced earlier are included in the experiments: Wachter [@wachter2017counterfactual], REVISE [@joshi2019towards], CLUE [@antoran2020getting], DiCE [@mothilal2020explaining] and Greedy [@schut2021generating]. In addition, we introduce two new generators in Section \@ref(mitigate) that directly address the issue of endogenous domain and model shifts. We also test to what extent it may be beneficial to combine ideas underlying the various generators.

>>>>

This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\@ref(empirical-2)). For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section. A few high-level details upfront: each experiment is run for a total of $T=50$ rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so $B_t=0.05 * N_t^{\mathcal{D}_0}$^[As mentioned in the previous section, we end up providing recourse to a total of $\approx50\%$ by the end of round $T=50$.]. All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment. Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model. Evaluation metrics are computed and stored every 10 rounds. To account for noise, each individual experiment is repeat five times.

## $M$---Classifiers and Generative Models {#empirical-classifiers}

For each dataset and generator, we look at three different types of classifiers, all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier---**Logistic Regression**---implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier. We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty. They are also the model of choice in @schut2021generating. The actual neural network architectures are kept simple (top half of Table \@ref(tab:architecture)), since we are only marginally concerned with achieving good initial classifier performance.

@@

## $\mathcal{D}$---Data {#empirical-data}

We have chosen to work with both synthetic and real-world datasets. Using synthetic data allows us to impose distributional properties that may affect the resulting recourse dynamics. Following @upadhyay2021robust, we generate synthetic data in $\mathbb{R}^2$ to also allow for a visual interpretation of the results. Real-world data is used in order to assess if endogenous dynamics also occur in higher-dimensional settings.

@@

We use four synthetic binary classification datasets consisting of 1000 samples each: **Overlapping**, **Linearly Separable**, **Circles** and **Moons**. The datasets are presented in Figure \@ref(fig:synthetic-data).

@@

Ex-ante we expect to see that by construction, Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary as we saw in Figure \@ref(fig:poc). Thus, the choice of a black-box model may have an impact on the paths of the recourse. For generators that use latent space search (REVISE @joshi2019realistic, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples. Of course, this expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty. Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class^[As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated. We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.]. In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concerned with generating what we have defined as meaningful counterfactuals.

@@

We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification. Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@kaggle2011give], originally consisting of 250,000 instances with 11 numerical attributes. Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable whether credit card clients default on their payment. In its raw form it consists of 23 explanatory variables: 4 categorical features relating to demographic attributes^[These have been omitted from the analysis. See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding. Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019realistic and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.

As a third dataset we include the **California Housing** dataset derived from the 1990 U.S. census and sourced through scikit-learn [@pedregosa2011scikitlearn, @pace1997sparse]. It consists of 8 continuous features that can be used to predict the median house price for California districts. The continuous outcome variable is binarized as $\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}$ indicating whether or not the median house price of a given district is above or below the median of all districts. While we have not seen this dataset used in the previous literature on AR, others have used the Boston Housing dataset in a similar fashion (see for example @schut2021generating). While we initially also conducted experiments on that dataset, we eventually discarded this dataset due to surrounding ethical concerns [@carlisle2019racist].

@@

## $G$---Generators

All generators introduced earlier are included in the experiments: Wachter [@wachter2017counterfactual], REVISE [@joshi2019realistic], CLUE [@antoran2020getting], DiCE [@mothilal2020explaining] and Greedy [@schut2021generating]. In addition, we introduce two new generators in Section \@ref(mitigate) that directly address the issue of endogenous domain and model shifts. We also test to what extent it may be beneficial to combine ideas underlying the various generators.

### 2022-08-19 paper/sections/empirical_2.rmd
incorporated Cynthia's comments

Below we first present our main experimental findings regarding these questions. We conclude this section with a brief recap providing answers to all of these questions.

@@

The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for Latent Space generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average. Related to this the bottom two rows indicate that the retrained classifiers disagree with their initial counterparts on the classification of up to nearly 25 percent of the individuals. We also note that the final classifiers are more decisive, although as we noted earlier this may to some extent just be a byproduct of retraining the model throughout the course of the experiment.

@@

Turning to the real-world data we will go through the findings presented in Figure \@ref(fig:real) where each column corresponds to one of the three data sets. The results shown here are for the deep ensemble, which once again largely resemble those for the MLP. Starting from the top row, perhaps somewhat surprisingly we find no substantial domain shifts. While Latent Space search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.

>>>>

Below, we first present our main experimental findings regarding these questions. We conclude this section with a brief recap providing answers to all of these questions.

@@

The same broad pattern also emerges in the third row: we observe the smallest deterioration in model performance for Latent Space generators, albeit we still find a reduction in the F-Score of around 5-10 percentage points on average. Related to this, the bottom two rows indicate that the retrained classifiers disagree with their initial counterparts on the classification of up to nearly 25 percent of the individuals. We also note that the final classifiers are more decisive, although as we noted earlier this may to some extent just be a byproduct of retraining the model throughout the course of the experiment.

@@

Turning to the real-world data we will go through the findings presented in Figure \@ref(fig:real), where each column corresponds to one of the three data sets. The results shown here are for the deep ensemble, which once again largely resemble those for the MLP. Starting from the top row, perhaps somewhat surprisingly we find no substantial domain shifts. While Latent Space search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.

### 2022-08-19 paper/sections/introduction.rmd
incorporated Cynthia's comments

Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences. While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of problem in the context of data-driven decision-making. They are essentially **black boxes**: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools. Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@o2016weapons]. The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models. These two sub-domains can be broadly referred to as **interpretable AI** and **explainable AI** (XAI), respectively.

Among the approaches to XAI that have recently grown in popularity are **Counterfactual Explanations** (CE). They explain how inputs into a model need to change for it to produce different outputs. Counterfactual Explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals who face adverse outcomes. An example relevant to the Social Sciences is consumer credit: in this context, AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system. A meaningful recourse recommendation for a denied applicant could be: *"If your net savings rate had been 10% of your monthly income instead of the actual 8%, your application would have been successful. See if you can temporarily cut down on consumption."* In the remainder of this paper we will use both terminologies - recourse and counterfactual - interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.

@@

We think that these types of endogenous dynamics may be problematic and warrant our attention. From a purely technical we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse. Secondly, we observe in Figure \@ref(fig:poc) that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter. We would argue that in some practical applications this can be expected to generate costs for the involved stakeholders. To follow our argument, consider the following two examples:

::: {.example #consumer name="Consumer Credit"}
Suppose Figure \@ref(fig:poc) relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their credit worthiness. Assume that the two features are actually meaningful in the sense that creditworthiness increases in the south-east direction. Then we can think of the outcome in panel (d) as representing a situation where the bank supplies credit to more borrowers (orange), but these borrowers are on average less creditworthy and more of them can be expected to default on their loan. This represents a cost to the retail bank.
:::

@@

To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR. Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Section \@ref(related) and \@ref(method)). Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\@ref(method-2)). Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting (\@ref(empirical) and \@ref(empirical-2)). Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \@ref(mitigate)). Finally, we discuss our findings in the broader context of the literature in Section \@ref(discussion), before pointing to some of the limitations of our work as well as avenues for future research in Section \@ref(limit). Section \@ref(conclusion) concludes.

>>>>

Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences. While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise to a new type of problem in the context of data-driven decision-making. They are essentially **black boxes**: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools. Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@oneil2016weapons]. The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models. These two sub-domains can be broadly referred to as **interpretable AI** and **explainable AI** (XAI), respectively.

Among the approaches to XAI that have recently grown in popularity are **Counterfactual Explanations** (CE). They explain how inputs into a model need to change for it to produce different outputs. Counterfactual Explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals who face adverse outcomes. An example relevant to the Social Sciences is consumer credit: in this context, AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system. A meaningful recourse recommendation for a denied applicant could be: *"If your net savings rate had been 10% of your monthly income instead of the actual 8%, your application would have been successful. See if you can temporarily cut down on consumption."* In the remainder of this paper we will use both terminologies---recourse and counterfactual---interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.

@@

We think that these types of endogenous dynamics may be problematic and warrant our attention. From a purely technical perspective we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse. Secondly, we observe in Figure \@ref(fig:poc) that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter. We would argue that in some practical applications, this can be expected to generate costs for the involved stakeholders. To follow our argument, consider the following two examples:

::: {.example #consumer name="Consumer Credit"}
Suppose Figure \@ref(fig:poc) relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their creditworthiness. Assume that the two features are actually meaningful in the sense that creditworthiness increases in the south-east direction. Then we can think of the outcome in panel (d) as representing a situation where the bank supplies credit to more borrowers (orange), but these borrowers are on average less creditworthy and more of them can be expected to default on their loan. This represents a cost to the retail bank.
:::

@@

To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR. Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Section \@ref(related) and \@ref(method)). Secondly, we introduce an experimental framework extending previous work by @altmeyer2022counterfactualexplanations, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (\@ref(method-2)). Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019realistic, @mothilal2020explaining and @antoran2020getting (\@ref(empirical) and \@ref(empirical-2)). Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \@ref(mitigate)). Finally, we discuss our findings in the broader context of the literature in Section \@ref(discussion), before pointing to some of the limitations of our work as well as avenues for future research in Section \@ref(limit). Section \@ref(conclusion) concludes.

### 2022-08-19 paper/sections/limitations.rmd
incorporated Cynthia's comments

Perhaps the most crucial shortcoming of our work is that we merely point out that there exists a trade-off between private costs to the individual and external costs to the collective of stakeholders. We fall short of providing any definite answers as to how that trade-off may be resolved in practice. The mitigation strategies we have proposed here provide a good starting point, but they are ad-hoc, mechanical extensions of the existing AR framework. An interesting idea to explore in future work could be the potential for Pareto optimal Algorithmic Recourse, that is a collective recourse outcome in which no single individual can be made better off, without make at least one other individual worse off. This type of work would be interdisciplinary and could help to formalize some of the concepts presented in this work. For the time being, we suggest that the trade-off can be resolved on a case-by-case basis.

@@

The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion. In practice, models are in fact updated on a regular basis [@upadhyay2021towards]. We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time $t=0$. That being said, our experimental design is a vast over-simplification of potential real-world scenarios. In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in @upadhyay2021towards. We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class. Relating this back to the consumer credit example, we assume that the would-be borrowers are always willing and able to implement recourse and the bank is always willing to provide credit as would-be borrowers move across the decision boundary. In practice it is doubtful that agents behave according to such simple rules. Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).

@@

For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks. While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice - both black-box and glass-box. Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (@borisov2021deep, @grinsztajn2022tree).

>>>>

Perhaps the most crucial shortcoming of our work is that we merely point out that there exists a trade-off between private costs to the individual and external costs to the collective of stakeholders. We fall short of providing any definite answers as to how that trade-off may be resolved in practice. The mitigation strategies we have proposed here provide a good starting point, but they are ad-hoc, mechanical extensions of the existing AR framework. An interesting idea to explore in future work could be the potential for Pareto optimal Algorithmic Recourse, that is, a collective recourse outcome in which no single individual can be made better off, without making at least one other individual worse off. This type of work would be interdisciplinary and could help to formalize some of the concepts presented in this work.

@@

The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion. In practice, models are in fact updated on a regular basis [@upadhyay2021robust]. We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time $t=0$. That being said, our experimental design is a vast over-simplification of potential real-world scenarios. In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in @upadhyay2021robust. We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class. Relating this back to the consumer credit example, we assume that the would-be borrowers are always willing and able to implement recourse and the bank is always willing to provide credit as would-be borrowers move across the decision boundary. In practice it is doubtful that agents behave according to such simple rules. Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).

@@

For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks. While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice. Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (@borisov2021deep, @grinsztajn2022why).
